Spatial Reasoning and Data Displays

### 2015-01-11 Paper/Draft1/TestingVisualAptitude.Rnw
Changes to main body of paper- intro, conclusion, moving sample tests to appendix, etc.

Lineups provide a quantitative measurement of the effectiveness of a particular plot: if participants consistently identify the target plot rather than the randomly-generated distractors, the plot effectively shows the difference between real data and random noise. 
This removes much of the subjectivity from user evaluations of display effectiveness, and the procedure is simple enough that it does not generally require participants to be very familiar with data-based graphics.

% 
% \comment{`While previous research' - which research are you referring to? 
% Yifan's paper with the eyetracking study has looked at the spatial connection \cite{zhao2013mind}. But is that what you mean?
% 
% `required by the lineup task' - that makes the lineup task strangely active. Could you rephrase that in terms of the participant?
% }

While previous research~\cite{lowrie2007solving,mayer1994whom} has examined the link between certain types of graphical perception and spatial skills, it is important to identify any additional visual skills participants utilize to complete the lineup task, as well as better understand demographic characteristics (math education, research experience, age, gender) which may impact performance \cite{humanfactorslineups}. 
% Exploring the participant characteristics which contribute to successful lineup evaluation allows us to better understand the lineup task and potential sources of variability in the study results.

\comment{good discussion, but this is the first time that you mention `our experiment' - include a sentence here, to say what this experiment is about. also: I think we should only say `study' or `experiment', and I'd prefer study, because experiment sounds like it could go BOOM. I don't want to give that impression. }

Our experiment compares lineup performance with established tests, such as the visual search task (VST), paper folding test, card rotation test, and figure classification test. 
The VST measures visual search speed\cite{goldstein1973validity}, the paper folding and card rotation tests measures spatial manipulation ability, and the figure classification test measures inductive reasoning\cite{ekstrom1976manual}; 
all of these skills are at least peripherally recruited during the lineup task, but some may dominate in predicting performance on the lineup task. 
We hope to facilitate comparison of the lineup task to known cognitive tests, inform the design of future studies, and better understand the perception of statistical lineups.

% 
% \comment{This is going into the right direction, but a few pieces are still missing, 
% you need to still say what this paper is doing: we do a study to compare lineup performance with established tests, also go into a bit of discussion of the tests.}
% \comment{Here is a paragraph missing, that connects the existing literature to what we are doing in this paper. 
% In particular, we need to answer the questions: 
% 
% - why are we interested in evaluating which skills are necessary to evaluate graphical displays (this has been implicitly answered by saying that cognitive load was reduced, so we free up resources and make graphics a better tool for more people.
% 
% - what are the limits of the studies that you have cited, ie is there something that we can do better?
% 
% - what advantage do lineups bring to the table?
% }
% 
% 
% \comment{How do perceptual principles apply to statistical graphics (data displays)? - Lineups help us to isolate different parts of displays and allow in-built quantification of results.}
% 
% 
% \comment{Why are lineups variable? - Individual variability is big, we want to investigate which abilities individuals help the most in evaluating. - for this reason we need to motivate why lineups are useful. }
% 
% 
% \comment{We examine 'some demographic characteristics' is a bit lukewarm. I think once the missing paragraph is there, you can sell the rest of the paper a bit more passionately.}

\comment{and in the last paragraph of the intro: lay out the roadmap for the rest of the paper. In section \ref{sec:methods} we introduce ... }

@@

\newtext{In a lineup test, the plot of the data is placed randomly among a set of, generally 19, distractor plots (or {\it null plots}) that are rendered from data generated by a model, without a signal (a null model). This sheet of charts is then shown to human observers, who are asked to identify the display that is ``the most different". If observers identify the plot drawn from the actual data, this can be reasonably taken as evidence that the data it shows is different from the data of other plots.
Let $X$ be the number of observers (out of $n$) who identify the data plot from the lineup. Under the null hypothesis that the data plot is not different from the other plots, $X$ has approximately a Binomial distribution \cite{wickham2010graphical, majumder2013validation}. If $k$ of the observers identify the data plot from the lineup, the probability $P(X \ge k)$ is the $p$-value of the corresponding visual test.

This aspect of lineups provides an important tool in the evaluation of graphical findings. Similarly, lineups allow for an objective evaluation of an individual's visual ability based on the general's population performance on a lineup.

}

\comment{Include how an individual's performance is evaluated. }

@@

The \textbf{visual searching task} (VST), shown in figure~\ref{fig:VST}, is designed to test a participant's ability to find a target stimulus in a field of distractors, thus making the visual search task similar in concept to lineups. Historically, visual search has been used as a measure of brain damage \cite{goldstein1973validity,demita1981validity,moerland1986neuropsychological}; however, similar tasks have been used to measure cognitive performance in a variety of situations, for example under the influence of drugs in~\cite{anderson1983interactive}. The similarity to the lineup protocol as well as the simplicity of the test and its' lack of color justify the slight deviation from forms of visual search tasks typically used in normal populations. 
\begin{figure}[htp]\centering
\includegraphics[width=\linewidth]{VisualSearch}
\caption[Visual Search Task]{Visual Search Task (VST). Participants are instructed to find the plot numbered 1-24 which matches the plot labeled ``Target". Participants will complete up to 25 of these tasks in 5 minutes.}\label{fig:VST}
\end{figure}

@@

\comment{Let's put figure \ref{fig:tests} into the appendix - that'll save us quite a bit of space. }
 \begin{figure*}[ht]
  \centering
   \subfloat[Figure Classification Task. Participants classify each figure in the second row as belonging to one of the groups above. Participants complete up to 14 of these tasks (each consisting of 8 figures to classify) in 8 minutes.\label{fig:figureclassification}]{%
  \includegraphics[width=.75\linewidth]{figureclassification}
}
\hfill
    \subfloat[Card Rotation Task. Participants mark each figure on the right hand side as either the same as or different than the figure on the left hand side of the dividing line. Participants complete up to 20 of these tasks (each consisting of 8 figures) in 6 minutes.\label{fig:cardrotation}]{%
\includegraphics[width=.75\linewidth]{cardrotation}
    }
    \hfill
    \subfloat[Paper Folding Task. Participants are instructed to pick the figure matching the sequence of steps shown on the left-hand side. Participants  complete up to 20 of these tasks in 6 minutes.\label{fig:paperfolding}]{%
\includegraphics[width=.75\linewidth]{paperfolding}
   }
    \caption{Visuospatial tests}
    \label{fig:tests}
  \end{figure*}

@@

\section{Results}
Results are based on an evaluation of \Sexpr{nrow(ans)} undergraduate students at Iowa State University. \Sexpr{round(sum(ans.summary$stem)/nrow(ans.summary)*100)}\% of the participants were in STEM fields, the others were distributed relatively evenly between agriculture, business, and the social sciences. Students were evenly distributed by gender, and were between 18 and 24 years of age with only one exception. This is reasonably representative\footnote{\url{http://www.registrar.iastate.edu/sites/default/files/uploads/stats/university/F14summary.pdf}} of the university as a whole; in the fall 2014 semester, 26\% of students were associated with the college of engineering, 24\% were associated with the college of liberal arts and sciences, 15\% were associated with the college of human sciences, 7\% with the college of design, 13\% with the business school, and 15\% with the school of agriculture.

@@

\comment{First: what is shown in figure 8? THis is a non traditional plot, so you need to explain it first.}
From figure~\ref{fig:lineupblockpca} we see that again, PC1 accounts for general visual aptitude. 
% PC2 emphasizes the relationship between figure classification and lineup task 1 (and to a lesser extent, on lineup task 2). PC3 differentiates the lineup task 3 from the visual search task; other tasks do not impact PC3 to a meaningful degree. PC4 separates lineup test 2 from the paper folding and card rotation tests and emphasizes the overlap with the visual search test. PC5 separates lineup test 2 and the figure classification test from the overlap between the other lineup tests and the visual search test. PC6 and PC7 account for a joint total of less than 10\% of the overall variability, and do not display any distinct patterns in the loadings.
%
PC2 emphasizes the overlapping variation in performance on lineup test 1 and the figure classification test. PC3 emphasizes the additional variation in performance on lineup test 3 and the visual search test, while PC4 emphasizes the extra variability in performance on lineup test 2 and the paper folding test. All three lineups, plus the figure classification, paper folding, and visual search tests contribute to PC5. PC6 and PC7 jointly account for less than 10\% of the variance in the data and do not display any distinct patterns in the loadings. 
%\Sexpr{as.numeric(round(100*importance[6], 1))}\% and \Sexpr{as.numeric(round(100*importance[7], 1))}\% of the variance, respectively; PC6 seems to differentiate the first two lineup tasks from the overlap between the third lineup task and the figure classification task.

@@

\section{Discussion and Conclusions}
\comment{These read like the conclusions from earlier - video games aren't that prominently displayed anymore unless you are referring to figure 5. But if so, I would be cautious about summarising STEM major as 'mathematical training' - use the same names for the variables as in figure 5 and make sure to keep the order, i.e. gender is more impotant than video game hours. also make sure, that to mention that this is on a univariate level and doesn't take the correlation between the variables into account. }

Performance on lineups is strongly related to performance on tests of visual ability; however, this relationship is mediated by factors such as mathematical training, hours of video games played per week, and gender.
In addition, many facets of intelligence are highly correlated; those who score higher on general aptitude tests may score higher on tests of visual ability (and may also score higher on lineup tests).
Despite these caveats, we have demonstrated that the general lineup task is most closely related to a classification task, rather than tests of spatial ability.
This is an important verification of a tool that is useful for examining statistical graphics, as it emphasizes the idea that while the testing medium is graphical in nature, the task is in fact a classification task, where the viewer must determine the most important features of each plot and then identify which plot is different.

@@

% \section{Examining potential testing biases}\label{app:ratests}
% T-tests of results for Hillary and Stephanie: 
% <<ttest, echo=F, include=T, results='asis'>>=
% ans.summary$tester <- c(rep("Hillary", 18), rep("Stephanie", nrow(ans.summary)-18))
% card.rotation <- t.test(card_rot~tester, data=ans.summary)
% paper.folding <- t.test(folding~tester, data=ans.summary)
% fig.class <- t.test(fig_class~tester, data=ans.summary)
% lineup <- t.test(lineup~tester, data=ans.summary)
% 
% res <- ldply(ans.summary[,c("card_rot", "folding", "fig_class", "lineup")], function(i){
%   tmp <- t.test(i~ans.summary$tester)
%   data.frame(Hillary = as.numeric(tmp$estimate)[1], Stephanie = as.numeric(tmp$estimate)[2],
%              t = tmp$statistic, df=tmp$parameter, p=tmp$p.value, lb = tmp$conf.int[1], ub=tmp$conf.int[2])
% })
% 
% res$.id <- c("Card Rotation", "Paper Folding", "figure classification", "Lineups")
% names(res) <- c("Variable", "Mean.Hillary", "Mean.Stephanie", "t", "df", "p-value", "Diff.LB", "Diff.UB")
% 
% print(xtable(res), include.rownames = F, sanitize.text.function=sanitize)
% @

>>>>

Lineups provide a quantitative measurement of the effectiveness of a particular plot: if participants consistently identify the target plot rather than the randomly-generated distractors, the plot effectively shows the difference between real data and random noise. 
This removes much of the subjectivity from user evaluations of display effectiveness, and the procedure is simple enough that it does not generally require participants to be very familiar with data-based graphics. 
While previous research~\cite{lowrie2007solving,mayer1994whom} has examined the link between certain types of graphical perception and spatial skills, it is important to identify any additional visual skills participants utilize to complete the lineup task, as well as better understand demographic characteristics (math education, research experience, age, gender) which may impact performance \cite{humanfactorslineups}.

\newtext{This study is designed to compare lineup performace with visual aptitude and reasoning tests, examining the skills necessary to successfully evaluate lineups. }
We compare lineup performace to the visual search task (VST), paper folding test, card rotation test, and figure classification test. 
The VST measures visual search speed\cite{goldstein1973validity}, the paper folding and card rotation tests measures spatial manipulation ability, and the figure classification test measures inductive reasoning\cite{ekstrom1976manual}; 
all of these skills are at least peripherally recruited during the lineup task, but some may dominate in predicting performance on the lineup task. 
We hope to facilitate comparison of the lineup task to known cognitive tests, inform the design of future studies, and better understand the perception of statistical lineups.

\newtext{
In section \ref{sec:methods}, we introduce the tests used in this study and describe how tests are scored. In section \ref{sec:results}, we discuss the test scores, comparing established tests with previously validated results, examining demographic characteristics associated with test scores. We discuss multicollinearity in the data, and use principal components analysis and linear regression to draw some conclusions about the similarity between lineups and aptitude tests. Finally, in section \ref{sec:discussion}, we discuss the implications of this study for the lineup protocol, and future work which could extend these results. 
}

@@

In a lineup test, the plot of the data is placed randomly among a set of, generally 19, distractor plots (or {\it null plots}) that are rendered from data generated by a model, without a signal (a null model). This sheet of charts is then shown to human observers, who are asked to identify the display that is ``the most different". If observers identify the plot drawn from the actual data, this can be reasonably taken as evidence that the data it shows is different from the data of other plots.
Let $X$ be the number of observers (out of $n$) who identify the data plot from the lineup. Under the null hypothesis that the data plot is not different from the other plots, $X$ has approximately a Binomial distribution \cite{wickham2010graphical, majumder2013validation}. If $k$ of the observers identify the data plot from the lineup, the probability $P(X \ge k)$ is the $p$-value of the corresponding visual test.

This aspect of lineups provides an important tool in the evaluation of graphical findings, in that we can aggregate responses from many individuals to evaluate a single graphic. In addition, however, we can aggregate an individual's score over several lineups to determine how well single individuals perform on the lineup task; data from previous lineup studies suggests that there is significant variation in individual performance as well. This study has the potential to explain some of that variance - if visual skills are important contributors, then perhaps we are simply measuring differences in spatial reasoning ability.
\newtext{
In order to evaluate an individual's overall lineup score, we would subtract a fraction of a point for each question answered incorrectly, and add a full point for each question answered correctly, as shown in 
\ref{eq.scoring1}.

\begin{eqnarray}\label{eq.scoring1}
\# \text{correct answers } - 1/(19) \cdot \# \text{wrong answers}.
\end{eqnarray}

This scoring scheme is chosen so that if participants are guessing, the expected score is 0. 
}
% 
% \comment{Include how an individual's performance is evaluated. }

@@

The \textbf{visual searching task} (VST), shown in figure~\ref{fig:VST}, is designed to test a participant's ability to find a target stimulus in a field of distractors, thus making the visual search task similar in concept to lineups. Historically, visual search has been used as a measure of brain damage \cite{goldstein1973validity,demita1981validity,moerland1986neuropsychological}; however, similar tasks have been used to measure cognitive performance in a variety of situations, for example under the influence of drugs in~\cite{anderson1983interactive}. The similarity to the lineup protocol as well as the simplicity of the test and its' lack of color justify the slight deviation from forms of visual search tasks typically used in normal populations.

@@

@@

\section{Results}\label{sec:results}
Results are based on an evaluation of \Sexpr{nrow(ans)} undergraduate students at Iowa State University. \Sexpr{round(sum(ans.summary$stem)/nrow(ans.summary)*100)}\% of the participants were in STEM fields, the others were distributed relatively evenly between agriculture, business, and the social sciences. Students were evenly distributed by gender, and were between 18 and 24 years of age with only one exception. This is reasonably representative\footnote{\url{http://www.registrar.iastate.edu/sites/default/files/uploads/stats/university/F14summary.pdf}} of the university as a whole; in the fall 2014 semester, 26\% of students were associated with the college of engineering, 24\% were associated with the college of liberal arts and sciences, 15\% were associated with the college of human sciences, 7\% with the college of design, 13\% with the business school, and 15\% with the school of agriculture.

@@

\newtext{Figure \ref{fig:lineupblockpca} shows the relative contribution each input variable makes to each principal component, weighted by the importance of each PC. This allows us to consider the rotation matrix visually; for instance, we see that again, PC1 accounts for most of the variance and seems to represent general visual aptitude. }
% PC2 emphasizes the relationship between figure classification and lineup task 1 (and to a lesser extent, on lineup task 2). PC3 differentiates the lineup task 3 from the visual search task; other tasks do not impact PC3 to a meaningful degree. PC4 separates lineup test 2 from the paper folding and card rotation tests and emphasizes the overlap with the visual search test. PC5 separates lineup test 2 and the figure classification test from the overlap between the other lineup tests and the visual search test. PC6 and PC7 account for a joint total of less than 10\% of the overall variability, and do not display any distinct patterns in the loadings.
%
PC2 emphasizes the overlapping variation in performance on lineup test 1 and the figure classification test. PC3 emphasizes the additional variation in performance on lineup test 3 and the visual search test, while PC4 emphasizes the extra variability in performance on lineup test 2 and the paper folding test. All three lineups, plus the figure classification, paper folding, and visual search tests contribute to PC5. PC6 and PC7 jointly account for less than 10\% of the variance in the data and do not display any distinct patterns in the loadings. 
%\Sexpr{as.numeric(round(100*importance[6], 1))}\% and \Sexpr{as.numeric(round(100*importance[7], 1))}\% of the variance, respectively; PC6 seems to differentiate the first two lineup tasks from the overlap between the third lineup task and the figure classification task.

@@

\section{Discussion and Conclusions}\label{sec:discussion}
% \comment{These read like the conclusions from earlier - video games aren't that prominently displayed anymore unless you are referring to figure 5. But if so, I would be cautious about summarising STEM major as 'mathematical training' - use the same names for the variables as in figure 5 and make sure to keep the order, i.e. gender is more impotant than video game hours. also make sure, that to mention that this is on a univariate level and doesn't take the correlation between the variables into account. }
\newtext{
Performance on lineups is strongly related to performance on tests of visual ability; however, this relationship is mediated by demographic factors such as major (STEM or not) and completion of calculus I.
In addition to these demographic factors, many facets of intelligence are highly correlated; those who score higher on general aptitude tests may score higher on tests of visual ability (and may also score higher on lineup tests).
}
Despite these caveats, we have demonstrated that the general lineup task is most closely related to a classification task, rather than tests of spatial ability.
This is an important verification of a tool that is useful for examining statistical graphics, as it emphasizes the idea that while the testing medium is graphical in nature, the task is in fact a classification task, where the viewer must determine the most important features of each plot and then identify which plot is different.

@@

\section{Visuospatial Aptitude Tests}
The Visual Search Task (VST), shown in figure \ref{fig:VST}, is used to measure a person's ability to locate a target amid a field of distractors. 
\begin{figure}[htp]\centering
\includegraphics[width=\linewidth]{VisualSearch}
\caption[Visual Search Task]{Visual Search Task (VST). Participants are instructed to find the plot numbered 1-24 which matches the plot labeled ``Target". Participants will complete up to 25 of these tasks in 5 minutes.}\label{fig:VST}
\end{figure}

Figure \ref{fig:tests} shows samples from the figure classification task, the card rotation task, and the paper folding task. All three tasks are part of the Kit of Factor-Referenced Cognitive Tests \cite{ekstrom1976manual}. 
 \begin{figure*}[ht]
  \centering
   \subfloat[Figure Classification Task. Participants classify each figure in the second row as belonging to one of the groups above. Participants complete up to 14 of these tasks (each consisting of 8 figures to classify) in 8 minutes.\label{fig:figureclassification}]{%
  \includegraphics[width=.75\linewidth]{figureclassification}
}
\hfill
    \subfloat[Card Rotation Task. Participants mark each figure on the right hand side as either the same as or different than the figure on the left hand side of the dividing line. Participants complete up to 20 of these tasks (each consisting of 8 figures) in 6 minutes.\label{fig:cardrotation}]{%
\includegraphics[width=.75\linewidth]{cardrotation}
    }
    \hfill
    \subfloat[Paper Folding Task. Participants are instructed to pick the figure matching the sequence of steps shown on the left-hand side. Participants  complete up to 20 of these tasks in 6 minutes.\label{fig:paperfolding}]{%
\includegraphics[width=.75\linewidth]{paperfolding}
   }
    \caption{Visuospatial tests}
    \label{fig:tests}
  \end{figure*}
% \section{Examining potential testing biases}\label{app:ratests}
% T-tests of results for Hillary and Stephanie: 
% <<ttest, echo=F, include=T, results='asis'>>=
% ans.summary$tester <- c(rep("Hillary", 18), rep("Stephanie", nrow(ans.summary)-18))
% card.rotation <- t.test(card_rot~tester, data=ans.summary)
% paper.folding <- t.test(folding~tester, data=ans.summary)
% fig.class <- t.test(fig_class~tester, data=ans.summary)
% lineup <- t.test(lineup~tester, data=ans.summary)
% 
% res <- ldply(ans.summary[,c("card_rot", "folding", "fig_class", "lineup")], function(i){
%   tmp <- t.test(i~ans.summary$tester)
%   data.frame(Hillary = as.numeric(tmp$estimate)[1], Stephanie = as.numeric(tmp$estimate)[2],
%              t = tmp$statistic, df=tmp$parameter, p=tmp$p.value, lb = tmp$conf.int[1], ub=tmp$conf.int[2])
% })
% 
% res$.id <- c("Card Rotation", "Paper Folding", "figure classification", "Lineups")
% names(res) <- c("Variable", "Mean.Hillary", "Mean.Stephanie", "t", "df", "p-value", "Diff.LB", "Diff.UB")
% 
% print(xtable(res), include.rownames = F, sanitize.text.function=sanitize)
% @
